---
title: "Report"
output:
  pdf_document: default
  html_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```
## Executive summary
In this report I aim to answer the questions 

“Is an automatic or manual transmission better for MPG”

"Quantify the MPG difference between automatic and manual transmissions"

My findings are that the differences between manual/automatic transmission are better explained through other variables.

## Dataset

```{r cars}
data(mtcars)
head(mtcars)
```
### Dataset description

*Description sourced from the R documentation*

The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973--74 models).

| Var  	| Desc                                     	|
|------	|------------------------------------------	|
| mpg  	| Miles/(US) gallon                        	|
| cyl  	| Number of cylinders                      	|
| disp 	| Displacement (cu.in.)                    	|
| hp   	| Gross horsepower                         	|
| drat 	| Rear axle ratio                          	|
| wt   	| Weight (1000 lbs)                        	|
| qsec 	| 1/4 mile time                            	|
| vs   	| Engine (0 = V-shaped, 1 = straight)      	|
| am   	| Transmission (0 = automatic, 1 = manual) 	|
| gear 	| Number of forward gears                  	|
| carb 	| Number of carburetors                    	|

### Initial exploration
vs and am are categorical variables so I will convert the columns from numeric to factor. I will also do the same for gear, cyl, and carb. Though it could be argued that there are numeric variables, in this case I will treat them as categorical.

```{r asfactor}
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$gear <- as.factor(mtcars$gear)
mtcars$carb <- as.factor(mtcars$carb)
mtcars$vs <- factor(mtcars$vs, labels=c("V-shaped", "Straight"))
mtcars$am <- factor(mtcars$am, labels=c("Automatic", "Manual"))
```

I will also rename the columns to more readable names
```{r rename}
names(mtcars) <- c("mpg", "cylinders", "displacement", "horsepower", "rear_axle_ratio", "weight", "quarter_mile_time", "engine", "transmission", "gears", "carburetors")
```

```{r}
str(mtcars)
summary(mtcars)
```

No NA values are present, so no imputation needed.
```{r na}
sapply(mtcars, function(x) sum(is.na(x)))
```

```{r mpg_plot}
ggplot(mtcars, aes(x = mpg)) +
    geom_density() +
    labs(title = "Distribution of MPG", x = "MPG", y = "Density")
```

```{r mpg_vs_transmision}
ggplot(mtcars, aes(x = mpg, fill = transmission)) +
    geom_density(alpha = 0.5) +
    labs(title = "Distribution of MPG by Transmission", x = "MPG", y = "Density")
```

It appears that automatic cars in general have a small range of MPG values, centred at a lower MPG than that of manual transmission cars, though manual transmission also have a larger range of values. We can verify this using a T-test, using a confidence level of 95% and a null hypothesis that the mean MPG of automatic and manual transmission cars are the same.
```{r t_test}
t.test(mpg ~ transmission, data = mtcars, conf.level = 0.95, alternative = "less")
```
We have a p-value of 0.0006 thus we can reject the null hypothesis that the mean MPG of automatic and manual transmission cars are the same, and accept the alternative hypothesis that the mean MPG of automatic transmission cars is less than that of manual transmission cars.

However this may be explaining relationships between other variables and mpg, assuming they are linked to transmission.

## Regression
```{r regression_transmission}
model <- lm(mpg ~ transmission, data = mtcars)
summary(model)
```
By fitting a linear model from just the transmission to mpg it seems that moving from a transmission of 0 (automatic) to 1 (manual) increases the mpg by 7.245. However this is not a very good model as the R-squared value is only 0.3598, meaning that only 35.98% of the variance in mpg is explained by the transmission.

```{r residual_plot}
ggplot(model, aes(x = .fitted, y = .resid)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed") +
    xlab("Fitted Values") +
    ylab("Residuals")
```


```{r regression_all}
model <- lm(mpg ~ ., data=mtcars)
summary(model)
```
When including all the variables we can see that the changing the transmission only increases the mpg by 1.212, suggesting that most of the change in mpg is due to other factors than just manual/automatic. The high p value of 0.7 suggests that we cannot attribute the change in mpg to transmission.

In particular horsepower and weight seem to contribute the most to capturing the variance of mpg. This model has an R-squared of 89%, and it is clear from the residual plot that it is a much better predictor of mpg.
```{r residual_plot2}
ggplot(model, aes(x = .fitted, y = .resid)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed") +
    xlab("Fitted Values") +
    ylab("Residuals")
```

## Conclusion
In conclusion: manual transmission is better for MPG than automatic, switching to manual will increase MPG by approximately 1.212, but I have shown that this transmission is not a statistically signifcant contributor to change in mpg when taking other variables into account.
